Moment Distances from Robust Subspace for Botnet Detection

### 2019-05-12 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
changes during the qcon

\title{Higher Moments Distances from Robust Subspace for Botnet Detection}

@@

\begin{abstract}
Distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid botnet's attacks. This paper proposes an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations. Experimental results reveal that the proposed approach outperforms widely adopted algorithms and a method based on robust estimates and robust skewness for network anomaly detection.
\end{abstract}

\begin{keyword}
Network Anomaly Detection \sep Botnet Detection \sep Higher Moments \sep Robust Principal Component Analysis \sep Mahalanobis Distance
\end{keyword}

@@

Network anomaly detection problems are usually characterized by skewed and imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. Learning algorithms for imbalanced data has been a challenging research topic, considering that the fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. Some techniques can make the data discriminative for better results on imbalanced data: Data engineering (i.e. sampling \cite{he2008learning,gu2008botminer}, aggregation \cite{lakhina2005mining,gu2008botminer,callegari2011novel, garcia2014empirical, acarali2016survey}); Feature extraction (i.e. decomposition \cite{vaswani2018robust}, robust estimates \cite{zhou2017anomaly}).

@@

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some widely adopted algorithms assume a gaussian distributed data, however this assumption may not be observed in real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}. Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings show that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network anomaly detection. Therefore, learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and kurtosis are higher moments that can have their measurement as very sensitive for outlier detection, considering that the properties of the kurtosis has been used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

@@

% botnet detection
A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information ex-filtration. We refer to \cite{ahmed2016survey,moustafa2019holistic} for an overview of network attacks. The botmaster coordinate a botnet through a command and control (C\&C) channel where bots receive commands and synchronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and architectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

@@

% justificative for the dataset and present who else use it
Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. However, NSL-KDD data set maintain the limitations of the KDD 99 regarding volume and reproduction of network traffic similar to the real networks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The authors recommend that none of the botnet families used in the training and cross-validation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by robust subspace larning
The problem of PCA or subspace learning for outlier corrupted data is called robust PCA (RPCA) or robust subspace learning \cite{vaswani2018robust}, which can be defined as the decomposition of a given data matrix $\textbf{M}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning. Robust subspace learning has receiving a growing attention of researchers aiming the development of anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.
RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of location and scatter matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

@@

% feature engineering and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Lakhina \emph{et al.} show that the analysis of feature distributions leads to significant advances on sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, while the anomaly detection approach leads gains on automatic classification of anomalies via unsupervised learning. Goldberg and Shan \cite{goldberg2015importance} argue, based on experience and on published real-life examples of anomaly detectors in use at eBay, that processing monitored signals into features is a more fruitful for anomaly detection systems than devising complex statistical tests.

@@

Considering that learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}, higher moments can be used for outlier detection and discriminative analysis of skewed data. Skewness and the kurtosis are higher moments that can contribute to sensitive outlier detectors, taking into account that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection in skewed data sets \cite{hubert2009robustskewed}.

@@

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adopted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. Finally, We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

The data set originally contains the following features:  Start Time, End Time, Duration, Source IP Address, Source Port, Direction, Destination IP Address, Destination Port, State, SToS, Total of Packets and Total of Bytes.

@@

An Exploratory Data Analysis (EDA) reveals that distributions of original features of the CTU-13 are skewed and presents an overlapping between normal and anomalous flows. Following we present the Figure \ref{fig:fig01}, that presents the distribution of flow duration for scenario 12, and exemplifies the data skewness overlapping. We adopt a sampling method based on data aggregation for obtaining more discriminative features, in accordance to important approaches for network anomaly detection \cite{lakhina2005mining, callegari2011novel, garcia2014empirical, chandrashekar2014survey,acarali2016survey, vieira2017model}.

@@

@@

Figure \ref{fig:fig02} shows the distribution of average flow duration, for scenario 12, aggregated by windows of 0.15 and 2 seconds. In comparison to Figure \ref{fig:fig01}, that shows the distribution of flow duration for the same scenario, we can see that the feature based on the average is more discriminative.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_0_15s_12_avg_duration.png}
     \caption{Average flow duration for scenario 12, aggregated by 0.15 seconds.}
     \label{fig:fig02}
\end{figure}

The data is modeled into the matrix $\textbf{M}$ with rows as observations and columns the features.

@@

This section describes the proposed technique to detect synflood, fraggle and port scan, according to Figure \ref{fig:fig80}, which represents the overview of the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvelue for each $q$-th time frame. Next, in Subsection \ref{sec:prop_MOSSchemes}, we show how to apply the eiganvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     % \includegraphics[height=11cm, width=9cm]{results/figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:fig80}
\end{figure}

\subsection{Largest Eigenvalue by Time Frames}
\label{sec:prop_LargestEigenvaluebyTimeFrames}

We propose to extend Fast-MCD for computing robust moments (skewness and kurtosis), in order to improve the anomaly detection in a skewed and imbalanced data set of network traffic with normal and botnet traffic.

>>>>

\title{Moment Distances from Robust Subspace for Botnet Detection}

@@

\begin{abstract}
Distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid botnet's attacks. This paper proposes an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations. Experimental results reveal that the proposed approach outperforms widely adopted algorithms and a method based on robust skewness for a public data set of botnet attacks.
\end{abstract}

\begin{keyword}
Network Anomaly Detection \sep Botnet Detection \sep Moments \sep Robust Principal Component Analysis \sep Mahalanobis Distance
\end{keyword}

@@

Network anomaly detection problems are usually characterized by skewed and imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. Learning algorithms for imbalanced data has been a challenging research topic, considering that the fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. Some techniques can make the data discriminative for better results on the imbalanced learning problem: Data engineering (i.e. sampling \cite{he2008learning,gu2008botminer}, aggregation \cite{lakhina2005mining,gu2008botminer,callegari2011novel, garcia2014empirical, acarali2016survey}); Feature extraction (i.e. decomposition \cite{vaswani2018robust}, robust estimates \cite{zhou2017anomaly}).

@@

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some widely adopted algorithms assume a gaussian distributed data, however this assumption may not be observed in real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}. Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings show that the skeweness and heavy-tailed distributions can impact algorithms that rely on gaussian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network anomaly detection. Therefore, learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and kurtosis are higher moments that can have their measurement as very sensitive for outlier detection, considering that the properties of the kurtosis has been used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

@@

% botnet detection
A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information ex-filtration. We refer to \cite{ahmed2016survey} and \cite{moustafa2019holistic} for an overview of network attacks. The botmaster coordinate a botnet through a command and control (C\&C) channel where bots receive commands and synchronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and architectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

@@

% justification for the data set and present who else use it
Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. However, NSL-KDD data set maintain the limitations of the KDD 99 regarding volume and reproduction of network traffic similar to the real networks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, and a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real network. The authors recommend that none of the botnet families used in the training and cross-validation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called robust PCA (RPCA) or robust subspace learning \cite{vaswani2018robust}, which can be defined as the decomposition of a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

Robust subspace learning has receiving a growing attention of researchers aiming the development of anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

@@

% feature engineering and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Lakhina \emph{et al.} show that the analysis of feature distributions leads to significant advances on sensitive detection of a wide range of anomalies, augmenting detection by volume-based methods, while the anomaly detection approach leads gains on automatic classification of anomalies via unsupervised learning. Goldberg and Shan \cite{goldberg2015importance} argue, based on experience and on published real-life examples of anomaly detectors in use at eBay, that processing monitored signals into features is a more fruitful for anomaly detection systems than devising complex statistical tests.

@@

Considering that learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}, higher moments can be used for outlier detection and discriminative analysis of skewed data. The skewness and kurtosis are higher moments that can contribute to sensitive outlier detectors, taking into account that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection in skewed data sets \cite{hubert2009robustskewed}.

@@

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adopted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. Finally, We evaluate the accuracy of the proposed approach for botnet detection on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

The data set originally contains the following features by flows:  Start Time, End Time, Duration, Source IP Address, Source Port, Direction, Destination IP Address, Destination Port, State, Type of service from source to destination, Type of service from destination to source, Total of Packets and Total of Bytes.

@@

An Exploratory Data Analysis (EDA) reveals that distributions of original features of the CTU-13 are skewed and presents an overlapping between normal and anomalous flows. Following we present the Figure \ref{fig:fig01}, that presents the distribution of flow duration for scenario 12, and exemplifies the data skewness and overlapping.

@@

Figure \ref{fig:fig02} shows a boxplot of flow duration on scenario 12, for normal and anomalous classes, in order to highlight the overlapping between classes.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_boxplot_0_15s_12_avg_duration.png}
     \caption{Flow duration for scenario 12, aggregated by 0.15 seconds.}
     \label{fig:fig02}
\end{figure}

We adopt a sampling method based on data aggregation for obtaining more discriminative features, in accordance to important approaches for network anomaly detection \cite{lakhina2005mining, callegari2011novel, garcia2014empirical, chandrashekar2014survey,acarali2016survey, vieira2017model}.

@@

Figure \ref{fig:fig03} shows the distribution of average flow duration, for scenario 12, aggregated by windows of 0.15 seconds. In comparison to Figure \ref{fig:fig01}, that shows the distribution of flow duration for the same scenario, we can see that the feature based on the average flow duration is more discriminative and shows a better separation between normal and anomalous classes.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_0_15s_12_avg_duration.png}
     \caption{Average flow duration for scenario 12, aggregated by 0.15 seconds.}
     \label{fig:fig03}
\end{figure}

The data is modeled into the matrix $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ with $n$ rows as observations and $p$ columns of features. Each scenario

@@

This section describes the proposed approach for botnet detection by means of moment distances from robust estimates. We propose to learn a robust subspace $\textbf{L}$ from normal traffic $\textbf{X}_n$ for computing robust moments, i.e. the mean $\textbf{l}$, skewness $\textbf{s}$ and kurtosis $\textbf{k}$, in order to evaluate the distance $\textbf{d}$ of new observations $\textbf{X}_o$ of these moments and classify as anomalous the observations with the largest distances.

Initially, the data $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ is split into normal data $\textbf{X}_n$ for training from normal data, and for $\textbf{X}_v$ and $\textbf{X}_t$ for respectively cross-validation and testing from contaminated data

It was shown that the RPCA problem can be solved via convex optimization that minimizes a combination of the nuclear norm and the $l_1$-norm, and that the method of augmented Lagrange multipliers (ALM) has been adopted to solve this convex problem.

Robust PCA aims to recover a low-rank matrix $\textbf{L}$ from corrupted measurements $\textbf{M} = \textbf{L} + \textbf{S}$. This decomposition in low-rank and sparse matrices can be achieved by techniques such as Principal Component Pursuit method (PCP),[1] Stable PCP,[2] Quantized PCP,[3] Block based PCP,[4] and Local PCP.[5] Then, optimization methods are used such as the Augmented Lagrange Multiplier Method (ALM[6]), Alternating Direction Method (ADM[7]), Fast Alternating Minimization (FAM[8]) or Iteratively Reweighted Least Squares (IRLS [9][10][11]).

Therefore we have $\textbf{L}$...
